---
title: "Getting started with chromadyn"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This walks through the bundled demo and then through adapting it to your own
data. Run it from the repository root.

## The 60-second version

```bash
pixi install
pixi run demo
```

That writes `results/report.html`. Everything below explains what it did.

## The inputs

The demo ships three files in `demo/`, derived from a published CD8+ T cell
ATAC timecourse. `demo/PROVENANCE.md` records exactly how, including the three
judgement calls involved.

```{r inputs}
ss <- read.delim("demo/samplesheet.tsv")
ss
```

Nine libraries, one arm, four timepoints in days. `time` is a number; its
units live in `config/demo.yaml` as `time_unit: days` and are never parsed
from the value.

```{r counts}
cts <- read.delim("demo/counts.tsv", row.names = 1, check.names = FALSE)
dim(cts)
cts[1:3, 1:4]
```

Raw integer counts, 5,000 peaks. The row names match the name column of
`demo/features.bed`, which is what switches on region mode.

## Running it

```bash
pixi run --frozen snakemake --configfile config/demo.yaml -j 4
```

Nine rules run: validate, prefilter, transform, qc, then per arm
differential, select_dynamic, cluster, superclusters, figures and export_bed,
then compare_arms, report and manifest.

## Reading the output

Start with validation. Sixteen checks run before anything expensive.

```{r validation}
v <- read.delim("results/qc/validation_report.tsv")
v[v$status != "PASS", ]   # empty on the demo
```

Then the PCA, before anything else. Replicates should sit on top of each
other and timepoints should separate. On the demo they trace a clean arc from
day 0 to day 8 with 93% of variance in two components. If yours does not,
nothing downstream is trustworthy.

Then the two gates:

```{r dynamic}
d <- read.delim(gzfile("results/differential/WT_results.tsv.gz"))
table(significant = d$padj < 0.01, big_enough = d$range >= 0.5)
```

A feature is dynamic only if both hold. See `docs/method.md` for why.

And the classes:

```{r classes}
read.delim("results/clusters/WT_supercluster_sizes.tsv")
```

```
supercluster_label      n
Decreasing           1435
Transient            1064
Sustained Increasing  320
Late Increasing       311
Transient Increasing  121
```

The per-feature table carries the class, the cluster it came from, how it was
assigned, and its coordinates:

```{r clusters}
cl <- read.delim("results/clusters/WT_clusters.tsv")
head(cl, 3)
```

## Adapting it to your data

Three things to get right, in order of how often they go wrong.

**Sample IDs must match exactly.** The `sample` column and the counts column
names, character for character. Quantifiers leave suffixes on; strip them
once, in the file, not by hand. A mismatch is a hard error that prints both
sides.

**`time` is numeric.** Put `4h` in `time_unit`, never in the column. A
character time column sorts lexicographically to `0, 120, 15, 240, 30, 60` and
turns every trajectory into noise; chromadyn asserts against this, but only
because it is the most destructive thing that can happen here.

**Scale `minc` to your feature count.** The default 50 suits roughly 14,000
clustered features. Use `minc: auto` if you are not sure.

Write a config that overrides only what differs:

```yaml
input:
  samplesheet: mydata/samplesheet.tsv
  counts: mydata/counts.tsv
  features: null            # gene mode
  baseline_group: shared
  time_unit: h
cluster:
  degpatterns:
    minc: auto
```

```bash
pixi run --frozen snakemake --configfile mydata/config.yaml -j 4
```

## Two arms sharing one baseline

The feature most likely to be why you are here. If two treatments share a
single set of untreated t=0 libraries, give those rows the sentinel group:

```
sample        group   time  replicate
unstim_r1     shared  0     r1
unstim_r2     shared  0     r2
egf_30m_r1    EGF     30    r1
hrg_30m_r1    HRG     30    r1
```

Each arm is then fitted separately against the same baseline, with no arm term
in either model, and `results/clusters/cross_arm_crosstab.tsv` tells you which
class a feature takes under each.

## When the class names come out numbered

If you get `Increasing-1`, `Increasing-2` rather than names, the shape rules
could not separate the pieces. That is information, not a failure. The log
prints each centroid's statistics and names the parameter to adjust; usually
`superclusters.split.cross`. `config/demo.yaml` carries a worked example, and
`docs/parameters.md` explains it.

## Debugging a single step

Any script runs outside Snakemake. Each starts with a block that builds an
equivalent `snakemake` object:

```bash
pixi run --frozen Rscript workflow/scripts/08_superclusters.R
```

Edit the `wildcards` and `configfile` in that block to point at your run.
