---
title: "timecourse-patterns on a public GEO timecourse"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Running timecourse-patterns on data you download yourself, in gene mode, on an RNA-seq
series rather than ATAC.

**Series:** [GSE227634](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE227634),
the RNA-seq arm of the SuperSeries the bundled demo's ATAC data comes from
(McDonald, Chick et al., *Immunity* 56(6):1303-1319.e5, 2023,
doi:10.1016/j.immuni.2023.05.005). Using the companion assay of the same
experiment is deliberate: it shows the same workflow on genes rather than
peaks, with nothing changed but the config.

## Status of this vignette

The **download is verified**: `vignettes/data/download_gse227634.sh` fetches
three files and checks them against SHA-256 digests recorded on 2026-09-22.
Run it and you will know immediately whether GEO still serves what this
vignette was written against.

The **analysis below has not been run end to end** in this repository, because
the merge step depends on decisions about the sample naming in the deposit
that are better made by someone looking at the data than baked in here. The
code is written out in full and the commands to validate it are given.

**This series has three timepoints, not four.** timecourse-patterns warns about that at
validation, and it is worth seeing why: with three points a transient rise
and a late rise cannot be told apart, so the split stage has little to work
with and you should expect coarse classes. It is a good illustration of the
warning doing its job, and a poor illustration of the five-class vocabulary.

## Download

```bash
bash vignettes/data/download_gse227634.sh
```

```
  ok  GSE227634_featureCounts.d3.mm39.vM27.txt.gz
  ok  GSE227634_featureCounts.d5.mm39.vM27.WT.txt.gz
  ok  GSE227634_featureCounts.d8.mm39.vM27.final.txt.gz
```

The deposit ships one featureCounts table per day rather than one matrix, so
the first job is to merge them.

## Merge into one counts matrix

```{r merge}
dir <- "vignettes/data/gse227634"
files <- c(d3 = "GSE227634_featureCounts.d3.mm39.vM27.txt.gz",
           d5 = "GSE227634_featureCounts.d5.mm39.vM27.WT.txt.gz",
           d8 = "GSE227634_featureCounts.d8.mm39.vM27.final.txt.gz")

read_fc <- function(path) {
  # Line 1 is the featureCounts command banner.
  x <- read.delim(path, skip = 1, check.names = FALSE, stringsAsFactors = FALSE)
  m <- as.matrix(x[, -seq_len(6), drop = FALSE])
  rownames(m) <- x$Geneid
  # Columns are BAM paths; keep the basename without the extension.
  colnames(m) <- sub("\\.bam$", "", basename(colnames(m)))
  m
}

mats <- lapply(file.path(dir, files), read_fc)

# Genes must be intersected, not assumed identical: the three tables were
# produced by separate featureCounts runs.
common <- Reduce(intersect, lapply(mats, rownames))
counts <- do.call(cbind, lapply(mats, function(m) m[common, , drop = FALSE]))
storage.mode(counts) <- "integer"
dim(counts)
```

Inspect the column names before going further. They encode genotype, cell
subset and replicate, and the encoding is not identical across the three
files. **This is the step to do by eye rather than trust.**

```{r inspect}
colnames(counts)
```

## Build the samplesheet

Keep the WT libraries, which are the ones present at every timepoint.

```{r samplesheet}
keep <- grep("WT", colnames(counts), value = TRUE)

samplesheet <- data.frame(
  sample = keep,
  group = "WT",
  time = c(d3 = 3, d5 = 5, d8 = 8)[sub("^.*?(d[358]).*$", "\\1", keep)],
  replicate = sprintf("r%d", ave(seq_along(keep), keep, FUN = seq_along)),
  stringsAsFactors = FALSE
)
stopifnot(!is.na(samplesheet$time), !anyDuplicated(samplesheet$sample))

dir.create("vignettes/data/run", showWarnings = FALSE)
write.table(data.frame(feature_id = rownames(counts), counts[, keep], check.names = FALSE),
            "vignettes/data/run/counts.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(samplesheet, "vignettes/data/run/samplesheet.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)
```

Derive `time` and `replicate` from the deposit's own metadata rather than from
a regex over filenames wherever you can. The regex above is a convenience for
a vignette; parsing structure out of filenames is exactly the practice the
samplesheet exists to replace, and `docs/method.md` explains what went wrong
when the source project did it.

## Configure

Gene mode: no coordinates, so BED export and annotation are skipped.

```yaml
# vignettes/data/run/config.yaml
input:
  samplesheet: vignettes/data/run/samplesheet.tsv
  counts: vignettes/data/run/counts.tsv
  features: null
  baseline_group: shared      # unused: this series has no untreated baseline
  time_unit: days

output:
  dir: vignettes/data/run/results

cluster:
  degpatterns:
    minc: auto                # genes are fewer than peaks; 50 would be too high

report:
  title: "timecourse-patterns: GSE227634 CD8+ T cell RNA timecourse"
```

## Run

```bash
pixi run --frozen snakemake --configfile vignettes/data/run/config.yaml -j 4
```

Expect validation to warn:

```
WARN time_count: only 3 timepoints. Shapes will be coarse and the transient
                 and late classes may not separate.
```

That is the point of including this series. Read the warning, then read
`results/clusters/<arm>_k_diagnostics.tsv` before believing any class count.

## Validating this vignette

```bash
bash vignettes/data/download_gse227634.sh            # checksums, verified
Rscript -e 'rmarkdown::render("vignettes/public-geo-timecourse.Rmd")'
```

If the checksums fail, GEO has reprocessed the deposit and the numbers here no
longer describe what you downloaded.
