---
title: "`r params$title`"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: kable
params:
  title: "timecourse-patterns results"
  results_dir: "results"
  arms: !r character()
  time_unit: "min"
  config: !r list()
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
R <- params$results_dir
arms <- params$arms
rd <- function(...) {
  p <- file.path(R, ...)
  if (file.exists(p)) utils::read.delim(p, check.names = FALSE, stringsAsFactors = FALSE) else NULL
}
fig <- function(name) {
  p <- file.path(R, "figures", paste0(name, ".png"))
  if (file.exists(p)) knitr::include_graphics(p) else invisible(NULL)
}
```

timecourse-patterns took a counts matrix over a timecourse and returned the features
that change over time, grouped by the shape of that change.

## Inputs

```{r inputs}
cfg <- params$config
knitr::kable(data.frame(
  setting = c("counts", "samplesheet", "features", "time unit", "arms",
              "transform", "differential", "FDR", "min_range",
              "cluster method", "cluster minc", "supercluster method"),
  value = c(cfg$input$counts, cfg$input$samplesheet,
            if (is.null(cfg$input$features)) "none (gene mode)" else cfg$input$features,
            params$time_unit, paste(arms, collapse = ", "),
            cfg$transform$method, cfg$differential$method,
            cfg$differential$fdr, cfg$differential$min_range,
            cfg$cluster$method, as.character(cfg$cluster$degpatterns$minc),
            cfg$superclusters$method)
))
```

## Validation

```{r validation}
v <- rd("qc", "validation_report.tsv")
if (!is.null(v)) {
  n <- table(factor(v$status, levels = c("PASS", "WARN", "FAIL")))
  cat(sprintf("%d checks: %d PASS, %d WARN, %d FAIL\n\n", nrow(v), n[["PASS"]],
              n[["WARN"]], n[["FAIL"]]))
  notable <- v[v$status != "PASS", ]
  if (nrow(notable)) knitr::kable(notable) else cat("Every check passed.\n")
}
```

```{r libsizes}
ls <- rd("qc", "library_sizes.tsv")
if (!is.null(ls)) knitr::kable(ls, caption = "Library sizes")
```

## Quality control

```{r qc-figs, out.width = "49%", fig.show = "hold"}
for (f in c("pca", "sample_correlation", "replicate_correlation")) {
  p <- file.path(R, "qc", paste0(f, ".png"))
  if (file.exists(p)) print(knitr::include_graphics(p))
}
```

Replicates should sit on top of each other and the timepoints should separate.
If they do not, nothing downstream is trustworthy.

## Dynamic features

```{r dynamic}
rows <- lapply(arms, function(a) {
  p <- file.path(R, "differential", paste0(a, "_results.tsv.gz"))
  if (!file.exists(p)) return(NULL)
  d <- utils::read.delim(gzfile(p), stringsAsFactors = FALSE)
  data.frame(arm = a, tested = nrow(d), dynamic = sum(as.logical(d$dynamic)),
             percent = round(100 * mean(as.logical(d$dynamic)), 1))
})
rows <- do.call(rbind, rows)
if (!is.null(rows)) knitr::kable(rows)
```

A feature is dynamic when it passes both gates: the significance test, and a
floor on the range of its per-timepoint means on the transformed scale. The
second gate is what stops a large experiment returning thousands of
statistically significant features whose total excursion is negligible.

## Trajectory classes

```{r superclusters, results = "asis"}
for (a in arms) {
  cat(sprintf("\n### %s\n\n", a))
  s <- rd("clusters", paste0(a, "_supercluster_sizes.tsv"))
  if (!is.null(s)) print(knitr::kable(s))
  cat("\n")
  for (f in c("_supercluster_ribbon", "_supercluster_traces", "_dendrogram",
              "_heatmap", "_degpatterns", "_dynamic_vs_static")) {
    p <- file.path(R, "figures", paste0(a, f, ".png"))
    if (file.exists(p)) cat(sprintf("![](%s)\n\n", p))
  }
}
```

### Choosing the number of classes

```{r kdiag}
kd <- do.call(rbind, lapply(arms, function(a) {
  d <- rd("clusters", paste0(a, "_k_diagnostics.tsv"))
  if (is.null(d) || !nrow(d)) NULL else cbind(arm = a, d)
}))
if (!is.null(kd) && nrow(kd)) {
  knitr::kable(kd, digits = 3,
               caption = "Silhouette and within-cluster sum of squares by k")
} else {
  cat("Too few clusters for a k diagnostic.\n")
}
```

Higher silhouette is better; the within-cluster sum of squares always falls
with k, so look for where it stops falling steeply. These describe cutting
the cluster dendrogram, which is a diagnostic here rather than the
assignment mechanism unless `superclusters.method` is `hclust`.

## Choosing the number of clusters

```{r clusterselection}
cs <- do.call(rbind, lapply(arms, function(a) {
  d <- rd("clusters", paste0(a, "_cluster_selection.tsv"))
  if (is.null(d)) NULL else cbind(arm = a, d)
}))
if (!is.null(cs)) {
  knitr::kable(cs)
}
```

How the cluster count was reached, which depends on `cluster.method`.
degPatterns cuts its divisive tree itself, so there is one row per arm.
k-means lists the mean silhouette width and within-cluster sum of squares of
every k it tried. The row marked `selected` is the one that was used.

## Cluster assignment quality

```{r assignqc}
aq <- do.call(rbind, lapply(arms, function(a) {
  d <- rd("clusters", paste0(a, "_cluster_assignment_qc.tsv"))
  if (is.null(d)) NULL else cbind(arm = a, d)
}))
if (!is.null(aq)) knitr::kable(aq)
```

## Across arms

```{r crosstab}
ct <- rd("clusters", "cross_arm_crosstab.tsv")
if (!is.null(ct)) {
  if (length(arms) == 1L) {
    cat("Single arm, so there is nothing to compare against.\n\n")
  }
  knitr::kable(utils::head(ct, 50))
}
```

## Reproducing this

Every threshold is in the config; the fully resolved copy, together with
package versions, every seed and SHA-256 checksums of the inputs, is written
to `run_manifest.json` beside this report.
